Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-09 response_to_reviewers_R1.Rmd

> Thank you for your feedback.

- If there is still space allowed in the abstract, I think you could consider explicitly
mentioning the three big ideas (expected value of a game state, win probability, and
measures of team strength). This would give readers a better idea of what is to come
and also allow people to find your article more easily if they are interested in one of
those specific topics.

> Thank you for this suggestion. We have added the three ideas explicitly to the abstract.

- In Section 2.1.2, I am a little confused as to what Carroll et. al. did to extend the
previous model of Carter and Machol, since both efforts seem to involve some sort of
linearity assumption.

> We have fixed the wording of this sentence. Instead of "extend", it now says "consider a similar approach".

- Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
one didn’t know the definition before looking it up (and as it turns out the concept is
quite simple)

@@

> We have added a sentence defining what an absorbing state is: "An absorbing state is a link in a Markov chain from which there are no possible transitions (i.e., it is the end of the chain)."

- In Section 2.1.3, you make the comment “they estimate that a better strategy would
be worth about 0.4 wins per season on average, a substantial amount in a season with
only 16 games.” This makes sense, but previously in Section 2.1.2 you mentioned
a potential difference of 3.8 wins per season, yet didn’t comment on the significant
magnitude of that effect. I’m not advocating for a particular change here, but the
authors could consider adding a small comment about the size of the effect Pelechrinis
et. al. estimated.
- At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
too conservative on 4th down decision making, but the effect may be overstated due
to the way in which to go yardage has been recorded in the past.” It’s unclear to me
whether the overstated effect is the one Lopez estimates in his paper, or if you are
referring to the estimated effects past studies which used less granular data.
- In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,
but it isn’t clear what those strategies are in reference to the previous discussion. Is
pulling the goalie optimal?

> We have added a clarification that coaches usually wait too long to pull their goalies.

- In the second formula of Section 4.1, the right hand side of the equation has a 2, but
it should be an α.

@@

- The authors have now made clearer the limitation in scope of this manuscript related to player evaluation. Perhaps, a statement regarding this can be added to the abstract.

> We have a statement to this effect to the abstract.

- I think that parts of the first paragraph of Section 5 regarding biometric data is a bit out of place. This could be moved to Section 7 (Opportunities). Likewise, Section 5 could be retitled something like‘Betting Data and Sports Analytics” or “Betting Data Analysis”.

> Thank you for this suggestion. We renamed Section 5: Sports betting market data and added this as a 4th "big idea". We moved the short discussion of the use of proprietary biometric data to the end of Section 7. We also added a short example of how betting lines could be used to compute implied team strengths.

### 2023-01-09 sports_analytics.Rmd
small citation fixes
added betting example; close #46

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect three big ideas that are common across multiple sports: the expected value of a game state, win probability, and measures of team strength.
  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources.
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  We also describe the use of non-performance data and opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
link-citations: yes
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: 
  rticles::arxiv_article:
    keep_md: yes
header-includes:
  - \usepackage{amsmath}
---

@@

In this paper, we explore three key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), and measures of team strength (Section \ref{sec:strength}). 
In each case, we define the concept mathematically, explain how it originated, and give examples of its applications in multiple sports. 
Our goal is to unify the conceptual threads, while doing some justice to the customizations necessary to make a metric meaningful in a particular sport. 
We include copious references to original works of scholarship.

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, we discuss the use of non-performance data (most notably, sports market betting data) in Section \ref{sec:betting}, and in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

# Non-performance data {#sec:betting}

Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
However, a growing body of literature is informed by non-performance data about sports. 
This research appears in fields like sports economics and kinesiology that are subfields of mature disciplines (a treatment of either is well beyond the scope of the this paper).
In addition to this body of research, we note that today, professional sports teams are collecting biometric data on their own players, and presumably using that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance. 
While this research may constitute "emerging methodologies," it will likely never be published, because each team will use it to their competitive advantage.

Nevertheless, in this Section we discuss concepts stemming from one notable, relatively new source of (semi-)public non-performance sports data: betting markets.
Since the 2018 United States Supreme Court decision in [*Murphy v. National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S. 
The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in at least two major ways.

@@

@@

>>>>

---
title: "Big Ideas in Sports Analytics and Statistical Tools for their Investigation"
authors:
  - name: Benjamin S. Baumer
#    thanks: Use footnote for providing further information about author (webpage, alternative address)---*not* for acknowledging funding agencies. Optional.
    department: Statistical & Data Sciences
    affiliation: Smith College
    location: Northampton, MA 01063
    email: bbaumer@smith.edu
    ORCID: 0000-0002-3279-0516
  - name: Gregory J. Matthews
    department: Mathematics and Statistics
    affiliation: Loyola University Chicago
    location: Chicago, IL 60660
    email: gmatthews1@luc.edu
    ORCID: 0000-0002-8413-5097
  - name: Quang Nguyen
    department: Statistics & Data Science
    affiliation: Carnegie Mellon University
    location: Pittsburgh, PA 15213
    email: nmquang@cmu.edu
abstract: |
  Sports analytics---broadly defined as the pursuit of improvement in athletic performance through the analysis of data---has expanded its footprint both in the professional sports industry and in academia over the past 30 years. 
  In this paper, we connect four big ideas that are common across multiple sports: the expected value of a game state, win probability, measures of team strength, and the use of sports betting market data.
  For each, we explore both the shared similarities and individual idiosyncracies of analytical approaches in each sport. 
  While our focus is on the concepts underlying each type of analysis, any implementation necessarily involves statistical methodologies, computational tools, and data sources.
  Where appropriate, we outline how data, models, tools, and knowledge of the sport combine to generate actionable insights. 
  We also describe opportunities to share analytical work, but omit an in-depth discussion of individual player evaluation as beyond our scope.
  This paper should serve as a useful overview for anyone becoming interested in the study of sports analytics. 
keywords:
  - sports analytics
  - R packages
  - sports data
  - pairwise comparisons
  - datasets
link-citations: yes
bibliography: [refs.bib, pkgs.bib]
biblio-style: unsrt   
# use apa for WIRE
csl: apa.csl
output: 
  rticles::arxiv_article:
    keep_md: yes
header-includes:
  - \usepackage{amsmath}
---

@@

In this paper, we explore four key ideas that have widespread applicability across many sports: the expected value of a game state (Section \ref{sec:ev}), win probability (Section \ref{sec:wp}), measures of team strength (Section \ref{sec:strength}), and the use of sports betting market data (Section \ref{sec:betting}). 
In each case, we define the concept mathematically, explain how it originated, and give examples of its applications in multiple sports. 
Our goal is to unify the conceptual threads, while doing some justice to the customizations necessary to make a metric meaningful in a particular sport. 
We include copious references to original works of scholarship.

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools (including a table of R packages) that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

# Sports betting market data {#sec:betting}

Most of the research in sports analytics is fueled by the analysis of data recorded from the outcome of sports contests.
However, a growing body of literature is informed by data from sports betting markets. 
Since the 2018 United States Supreme Court decision in [*Murphy v. National Collegiate Athletic Association*](https://en.wikipedia.org/wiki/Murphy_v._National_Collegiate_Athletic_Association), sports gambling has exploded in the U.S. 
The increasing interest in sports gambling has led to increasing interest in sports gambling *data*, and that data has proven useful to researchers in at least two major ways.

@@

## Example of how to use betting market data

To see how betting market data can be used to estimate team strengths, consider the betting odds posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals), posted on January 9, 2023 and shown in Figure \ref{tab:odds}. 
This is a futures market, because the actual NBA champion will not be determined until June 2023. 
The Boston Celtics are the favorite to win, with posted odds of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay out \$390 if the Celtics win it all. 
Conversely, six teams share the lowest odds at $+50000$.

```{r odds}
library(tidyverse)
odds <- read_csv("odds.csv")
odds |>
  mutate(
    prob = 100 / (100 + line),
    `Prob. Normalized` = prob / sum(prob)
  ) |>
  rename(Team = team, Line = line, `Prob.` = prob) |>
  mutate(Rank = row_number(), .before = Team) |>
  janitor::adorn_totals() |>
  slice(c(1:6, 25:31)) |>
  knitr::kable(
    booktabs = TRUE, 
    linesep = c('', '', '', '', '', '\\addlinespace'),
    digits = 3, 
    caption = "2023 NBA Championship line odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
```

These lines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
$$
  p_i = \frac{100}{100 + \ell_i} \,.
$$
The sum of those probabilities is greater than one---this is why the sportsbook makes money regardless of who wins the championship. 
However, the implied probabilities can be normalized by dividing by their sum to recover true probabilities of each team winning the championship. 
Many different researchers have shown that these normalized implied probabilities are accurate, unbiased, and efficient estimates of the true unknowable probabilities (see @Lopez2018 for discussion and an extensive list of references).

In this case, the FanDuel futures market suggests that the Celtics have a 16.3\% chance of winning the championship, while the Milwaukee Bucks have the second best chance, at 13.3\%. 
These implied probabilities can be used to fit various models for team strength, as described in Section \ref{sec:strength}.

@@

It is also worth noting that a significant, but unknown, proportion of the most innovative research is being conducted by professional sports teams. 
This research will likely never be published, because each team will use it to their competitive advantage. 
Part of what enables this research is better data. 
For example, professional sports teams can collect biometric data on their own players, and use that data to learn about how their workouts, sleep patterns, and diets impact their athletic performance. 
While this research may constitute "emerging methodologies," it unfortunately will take years, if at all, before the public benefits from it.
